Large-scale spatiotemporal variation in vital rates of an alpine bird

### 2023-12-05 Manuscript_Nater_etal.qmd
Set up manuscript structure with bullet points and text to recycle

## About this worknote

The purpose of this worknote is to introduce and showcase preliminary results from application of the open-population integrated distance-sampling model (IDSM) for willow ptarmigan to data from all of Norway. Eventually, this model extension and resulting insights into large-scale spatiotemporal variation in density, population growth rate, and vital rates will be the subject of a second scientific paper. This paper will be written and submitted in the second half of 2023, and this worknote serves as a temporary substitute for it in the reporting of project results to Miljødirektoratet in June 2022. As such, the worknote contains a general description of the extension of the model to analyse data from multiple areas simultaneously, visualizations of some preliminary results, and an introduction to a reproducible workflow that allows to easily and robustly re-run analyses as new data becomes available in future years.

## IDSM extension to multiple areas

Distance-sampling data on willow ptarmigan is collected on line transects not only in western Lierne, but across most of its Norwegian range (https://honsefugl.nina.no/). Taking the line transect datasets from Fjellstyrene, Statskog, and FeFo together, there is data for willow ptarmigan in 41 reporting areas (for a list, see [here]( https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp/blob/main/R/listAreas.R)). The model developed as part of this project and presented in @nilsen2023 was designed with the potential of extension to run on data from multiple areas simultaneously in mind, and this extension involves two steps in practice.

@@

## Preliminary results

In the following, we present some preliminary results from the model fit to data from all 41 reporting areas simultaneously. Results are presented in the format of pairs of maps. For each quantity/result, we present one map of the posterior median and one map of the standard deviation of the posterior distribution, where the latter serves as a measurement of uncertainty for the former.

### Population density

@@

### Population growth rate

@@

### Survival probability

@@

::: {#fig-pSurv layout-ncol="2"}

![Median survival estimates](Figures/pSurv_Median.png){#fig-pSurvA}

@@

### Recruitment rate

@@

### Effect of rodent occupancy

@@

![Graphical representation of the "targets pipeline" for the multi-area modelling setup. Upward facing triangles are functions, downward facing triangles are general options/arguments, circles are objects and outputs (="targets") created as part of the workflow. ](Figures/Workflow_layout.png){#fig-workflow width="424"}

## Data and code availability

@@

The auxiliary radio-telemetry data and all code, including the "targets pipeline" can be found in the project's repository on GitHub: [https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp](https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp).

## References

>>>>

# Abstract

# Introduction

# Methods

## Study species

-   Willow ptarmigan distribution, life history, and (historical) abundance

## Data collection, management, and preparation

### Line transect sampling

-   Setup of Hønsefugltakseringen monitoring programme (volunteer training, line transect sampling, spatio-temporal coverage, etc.)
-   Data & metadata recording via app
-   Quality control of data
-   Open access publishing of data (GBIF, LivingNorway)
-   Pre-modelling data wrangling

Distance-sampling data on willow ptarmigan is collected on line transects not only in western Lierne, but across most of its Norwegian range (https://honsefugl.nina.no/). Taking the line transect datasets from Fjellstyrene, Statskog, and FeFo together, there is data for willow ptarmigan in 41 reporting areas (for a list, see [here](https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp/blob/main/R/listAreas.R)).

### Radio-telemetry study in Lierne

-   Description of setup, data collection, temporal coverage, etc.
-   Pre-modelling data wrangling

## National-scale integrated model

### Integrated distance sampling model (IDSM) for willow ptarmigan

-   Introduction of integrated Distance sampling model (IDSM) from MS1

### Multi-area model extension

-   Addition of spatial dimension
-   Hyper-parameter distributions for spatial variation
-   Temporal variation (shared year RE and "residual" variation)

The model developed as part of this project and presented in @nilsen2023 was designed with the potential of extension to run on data from multiple areas simultaneously in mind, and this extension involves two steps in practice.

@@

Another approach to maximising information gain when jointly analysing data from different areas is explicit modelling of spatial and spatio-temporal variation.

### Model implementation

-   Bayesian, NIMBLE
-   Priors
-   Default samplers
-   Chains, iterations, burn-in, thinning, posterior samples

## Variance decomposition

-   How variance decomposition was done

![Graphical representation of the "targets pipeline" for the multi-area modelling setup. Upward facing triangles are functions, downward facing triangles are general options/arguments, circles are objects and outputs (="targets") created as part of the workflow.](Figures/Workflow_layout.png){#fig-workflow width="424"}

# Results

## Population density

@@

## Population growth rate

@@

## Survival probability

@@

::: {#fig-pSurv layout-ncol="2"}
![Median survival estimates](Figures/pSurv_Median.png){#fig-pSurvA}

@@

## Recruitment rate

@@

## Effects of rodent occupancy

@@

## Detection parameters

\[TBA\]

## Variance decomposition

\[TBA\]

# Discussion

\[TBA\]

# Acknowledgements

\[TBA\]

# Data and code availability

@@

The auxiliary radio-telemetry data and all code, including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.

# References
